Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes

### 2025-02-28 ms/ms.Rmd
addressed coauthor comments

  This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway. 

@@

  Then we trained random forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution. 

@@

  Still, weak relationships with remotely sensed variables make peat depth hard to predict overall. 
  Based on these findings, digital soil mapping can improve existing, broad-scale maps of peat depth, but local carbon stock assessment requires additional field sampling. 
  Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.

@@

    R code used in this study is available at <https://github.com/julienvollering/DSMdepth>. Depth measurements will be archived in a public repository upon publication.

@@

Peat soils are a terrestrial carbon stock of global importance. 
They store about 30 % of global soil carbon despite covering only 2--3 % of Earth's land [@xuPEATMAPRefiningEstimates2018; @friedlingsteinGlobalCarbonBudget2020; @unepGlobalPeatlandsAssessment2022]. 
Peatlands (areas with peat soil) are more carbon dense per square meter than any other ecosystem [@temminkRecoveringWetlandBiogeomorphic2022]. 
This makes them crucial to climate change mitigation. 
Intact peatlands sequester carbon and overall produce a negative temperature forcing [@joostenRolePeatlandsClimate2016]. 
When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions [@maGloballyRobustRelationship2022].

One of the keys to the areal density of peatland carbon stocks lies in the third dimension: their depth. 
Peat soils range from zero to over ten meters deep, so deep peats comprise a large volume. 
Their depth results from the accumulation of organic matter over thousands of years [@loiselDatabaseSynthesisNorthern2014; @joostenRolePeatlandsClimate2016]. 
In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat. 
There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposistion.

Peatlands are unevenly distributed globally, and in regions with high cover they are frequently converted to human land use [@unepGlobalPeatlandsAssessment2022].
They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils.
However, other land uses also displace peatlands. 
In Norway -- where flat lowland is relatively scarce but 9 % of the land area is peatland [@brynLandCoverNorway2018] -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
Since then, a larger proportion of peatland loss is driven by construction [@flagetMyrMellomLokal2024].

@@

Measuring peat depth on the ground by probing or ground-penetrating radar is straightforward, and a field survey can map a small area at low cost. 
However, surveying large areas is impractical when depth varies over short distances (e.g. < 1 m) -- as in most peatlands. 
A complementary approach from soil science is digital soil mapping (DSM). 
DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest. 
This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying patterns, especially through machine learning [@minasnyDigitalMappingPeatlands2019; @wadouxMachineLearningDigital2020].

The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped variables. 
For DSM to work, these relationships must be strong and consistent over the area of interest.
They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent than non-causal ones. 
The *scorpan* framework for DSM suggests seven predictor classes to explore: other soil properties, climate, organisms, relief (topography), parent material, age, and spatial position [@mcbratneyDigitalSoilMapping2003].

For peat depth, the most practical and widespread *scorpan* factors are relief and spatial position. 
Spatial position is unique because it is always known (with varying accuracy). 
However, the short range of spatial autocorrelation in peat depth limits its mapping value [@henglGenericFrameworkSpatial2004]. 
Relief, or topography, is widely and accurately mapped in digital terrain models (DTMs). 
For example, most of mainland Norway has at least one elevation measurement per square meter from airborne Light Detection and Ranging (LiDAR) surveys. 
Moreover, mechanisms of peat formation are linked to topography. 
For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation.

Studies have consistently shown relationships between peat depth and topography, although the specific patterns vary. 
One of the most robust relationships is a negative correlation between depth and terrain slope [e.g. @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @gatisMappingUplandPeat2019]. 
Peat depth also changes with elevation in many contexts, but with inconsistent directionality [e.g. @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @rudiyantoDigitalMappingCosteffective2016; @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024]. 
Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies [e.g. @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024]. 
Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic variables and the resolution of the peat depth analysis.

Another set of variables related to peat depth are measurements of natural radioactivity from the ground surface. 
Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in the earth's crust: potassium-40, uranium-238, and thorium-232 [@reinhardtGammaraySpectrometryVersatile2019]. 
These exist in bedrock (and mineral soils) and a peat overburden attenuates the radiation intensity at the surface. 
The degree of attenuation relates to properties of the overburden, especially composition and depth. 
Deep soil with high water content will attenuate radiation most [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019]. 
Thus, gamma-ray radiometrics integrate two *scorpan* factors: other soil properties and parent material [@mcbratneyDigitalSoilMapping2003].

@@

The effectiveness of DSM depends not only on the methods and data, but also on the characteristics of the mapping area. 
Norway may be instructive in this respect because its peatlands vary widely across climates and topographies. 
Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs [@joostenWiseUseMires2002; @lyngstadBeskrivelserAvTorvmassivenheter2023]. 
These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric variables will vary by landscape.

@@

The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration.
Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps *before* detailed investigation increases the option space for spatial planning and land management. 
For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work, but by that time, the route of the road is already set. 
With a digital peat depth map, planners could better compare routes and climate mitigation measures would have more leverage.
Therefore, soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations are.

Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at the landscape scale, with a view towards revising regional and national maps in Norway. 
Specifically, we: (1) quantify the accuracy of predictions from topographic and radiometric variables, (2) identify key predictive variables, and (3) compare these relationships across two different Norwegian landscapes. 
To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms. 
Where airborne radiometric data have previously been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation [e.g., @keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @siemonAirborneElectromagneticRadiometric2020]. 
Where machine learning algorithms have previously been used on airborne radiometric data, it has been to predict peat extent rather than depth [e.g., @olearyDigitalSoilMapping2022].

@@

At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage and accessibility (Fig. \@ref(fig:sites)b).
The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % mozonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (Geological Survey of Norway, 1:250 000 dataset).
The area within our delineation at Skrimfjella is classified as the landscape type *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
It is almost without human infrastructure, dominated by forest, and borders on a large nature reserve.
The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
In Norway's AR5 national land cover dataset ["areal resources in scale 1:5000", @ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as 'mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.

@@

We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \@ref(tab:preds)).
All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
From the radiometric surveys we simply used the four variables produced by the Geological Survey of Norway: ground concentration of Potassium, Thorium, Uranium, as well as total count.
From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables [@maxwellLandsurfaceParametersSpatial2022].
The categorical predictor was peat depth class, from a national map dataset.
We chose a spatial resolution of 10 m for our predictors, peat depth sample, and modelling.
We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other.
Complete descriptions of predictors follow below.

```{r preds, echo=FALSE, message=FALSE}
df <- readr::read_csv("tables/predictors.csv")
options(knitr.kable.NA = '')
knitr::kable(df, booktabs = FALSE, vline = "", linesep = "",
             col.names = c("Group", "Name", "Description"),
             caption = "Candidate predictors of peat depth.")
```

@@

The Geological Survey of Norway conducted and processed radiometric surveys over our study areas, as reported in Baranwal et al. [-@baranwalHelicopterborneMagneticElectromagnetic2013] and Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015].
They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the *terra* package in R.

The radiometric surveys were conduced with similar flight parameters (Table \ref{tab:radSurveys}).
Spectrometer count rates were calibrated to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
Data were processed following standard procedures outlined by the International Atomic Energy Association.
Processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
At Ørskogfjellet an additional convolution filter was added to smooth the gridded data.

@@

We used the *terra* R package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @dornikOptimalScalingPredictors2022; @newmanAssessingSpatiallyHeterogeneous2023].
We know that peat depth tends to vary at fine scales in Norway, which is why we chose 1 m and 10 m resolutions [@maxwellLandsurfaceParametersSpatial2022].
We also calculated the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
We calculated this index in SAGA GIS [v.9.3.2, Morphometry library, @conradSystemAutomatedGeoscientific2015] with default parameters (initial slope threshold = 16 %, lowness threshold = 0.4, upness threshold = 0.35, slope shape parameter = 4, elevation shape parameter = 3).

@@

The Depth-to-Water index [@murphyMappingWetlandsComparison2007] approximates a location's vertical height above the surface water feature that it is likely to drain towards.
It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature [eq. 5 in @murphyTopographicModellingSoil2009].
We calculated unitless slope from the 1 m DTM using the Whitebox software.
Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthwater2021].
This catchments area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
For each of the seven surface water layers, we derived Depth-to-Water using the *Distance Accumulation* tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.

@@

At Skrimfjella, we used the *eSample* function in the *iSDM* R package (v.1.0) to stratify our sample across elevation, slope, and potassium concentration.
This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and *eSample* returned 105 raster cells.

We also measured peatland occurrence (binary variable) in a separate sample at Skrimfjella.
We measured peatland occurrence because the AR5 dataset is known to underestimate peatland coverage [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
The 10 m cells were sampled from those outside AR5 mire and with slope < 20°. 
Following the same procedure, *eSample* returned 106 raster cells from this population.

@@

At Ørskogfjellet, we first determined a minimal sample size that would adequately capture the slope and radiometric properties (potassium, thorium, uranium, and total count) of the entire AR5 mire area [@sauretteDivergenceMetricsDetermining2023].
Specifically, we identified an elbow point in a curve of similarity between sample and population [@maloneMethodsImproveUtility2019].
For a sequence of sample sizes (50--500) [ten replicates each, drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback--Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
Then, we fitted a asymptotic regression of mean divergence on sample size, and found that the curve reached 95 % of the fitted asymptote at a sample size of 160.

To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the the *fields* package (v.14.2).
Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space [@brusSamplingDigitalSoil2019].
This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: slope and four radiometrics.

We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (with respect to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.

@@

We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system receivers.
We dampened the effect of outlying measurements by probing three times at each location [@parryEvaluatingApproachesEstimating2014], at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides. 
We used changes in resistance to indicate the base of the peat column.
Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact, like a buried rock.
Where the peat column was deeper than the extendable probe could be manually inserted and extracted by a pair of operators, we recorded a right-censored result (one at Skrimfjella, five at Ørskogfjellet).

>>>>

  This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in western and southeastern Norway. 

@@

  Then we trained Random Forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution. 

@@

  Still, weak relationships with remotely sensed variables made peat depth hard to predict overall.
  Based on these findings, we conclude that digital soil mapping can improve existing, broad-scale maps of peat depth in Norway, but highly localized carbon stock assessments are best made from field measurements.
  Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat lateral extent and depth.
  Together, these pathways promise more accurate landscape-scale carbon stock assessments and better-informed land management policies.

@@

    R code used in this study is available at <https://github.com/julienvollering/DSMdepth>. The depth measurements we used are archived at \doi{10.6073/pasta/6ce440152f693f2156bf5b692a2e7917} and follow data and metadata standards.

@@

Peat soils are a terrestrial carbon stock of global importance. 
They store 450-650 Gt of carbon, or about 30 % of global soil carbon, despite covering only 2--3 % of Earth's land [@xuPEATMAPRefiningEstimates2018; @friedlingsteinGlobalCarbonBudget2020; @unepGlobalPeatlandsAssessment2022]. 
Peatlands (areas with peat soil) are more carbon dense per square meter than any other ecosystem [@temminkRecoveringWetlandBiogeomorphic2022]. 
This makes them crucial to climate change mitigation. 
Intact peatlands sequester carbon and overall produce a negative temperature forcing [@joostenRolePeatlandsClimate2016]. 
When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions [@maGloballyRobustRelationship2022].

One of the keys to the areal density of peatland carbon stocks lies in the third dimension: their depth. 
Peat soils range from zero to over ten meters deep [@widyastutiPEATGRIDSMappingThickness2024], so deep peats comprise a large volume. 
Their depth results from the accumulation of organic matter over thousands of years [@loiselDatabaseSynthesisNorthern2014; @joostenRolePeatlandsClimate2016]. 
In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat. 
There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposition.

Peatlands are most common at high latitudes, and in regions with high cover they are frequently converted to human land use [@unepGlobalPeatlandsAssessment2022].
They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils.
However, other land uses also displace peatlands. 
In Norway -- where flat lowland is relatively scarce but upward of 9 % of the land area is peatland [@brynLandCoverNorway2018; @bakkestuenDelineationWetlandAreas2023] -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
Since then, a larger proportion of peatland loss is driven by construction [@flagetMyrMellomLokal2024].

@@

Measuring peat depth on the ground by probing or ground-penetrating radar is straightforward, and a field survey can map a small area at low cost. 
However, surveying large areas is impractical when depth varies widely over short distances (e.g. 10 m) -- as in many peatlands [@torppaStatisticalApproachDetermining2011; @proulx-mcinnisReviewGroundpenetratingRadar2013; @henrionMappingMonitoringPeatlands2024]. 
A complementary approach from soil science is digital soil mapping (DSM). 
DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest. 
This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying patterns, especially through machine learning [@minasnyDigitalMappingPeatlands2019; @wadouxMachineLearningDigital2020].

The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped predictors. 
For DSM to work, these relationships must be strong and consistent over the area of interest.
They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent than non-causal ones. 
The *scorpan* framework for DSM suggests seven predictor classes to explore: other soil properties, climate, organisms, relief (topography), parent material, age, and spatial position [@mcbratneyDigitalSoilMapping2003].

For peat depth, the most practical and widespread *scorpan* factors are relief and spatial position. 
Spatial position is unique because it is always known (with varying accuracy). 
However, the short range of spatial autocorrelation in peat depth limits its mapping value [@henglGenericFrameworkSpatial2004]. 
Relief, or topography, is widely and accurately mapped in digital terrain models (DTMs). 
For example, most of mainland Norway has at least one elevation measurement per square meter from airborne Light Detection and Ranging (LiDAR) surveys. 
Moreover, mechanisms of peat formation are linked to topography. 
For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation of peat.

Studies have consistently shown relationships between peat depth and topography, although the specific patterns vary. 
One of the most robust relationships is a negative correlation between depth and terrain slope [e.g. @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @gatisMappingUplandPeat2019]. 
Peat depth also changes with elevation in many contexts, but with inconsistent directionality [e.g. @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @rudiyantoDigitalMappingCosteffective2016; @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024]. 
Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies [e.g. @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024]. 
Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.

Another set of predictors related to peat depth are measurements of natural radioactivity from the ground surface. 
Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in the earth's crust: potassium-40, uranium-238, and thorium-232 [@reinhardtGammaraySpectrometryVersatile2019]. 
These exist in bedrock (and mineral soils) and a peat overburden attenuates the radiation intensity at the surface. 
The degree of attenuation relates to properties of the overburden, especially composition and depth. 
Deep soil with high water content will attenuate radiation most [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019]. 
Thus, gamma-ray radiometrics integrate two *scorpan* factors: other soil properties and parent material [@mcbratneyDigitalSoilMapping2003].

@@

The effectiveness of DSM depends not only on the methods and data, but also on the characteristics of the mapping area. 
Norway may be instructive in this respect because its peatlands vary widely across climates and topographies. 
Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs [@joostenWiseUseMires2002; @lyngstadBeskrivelserAvTorvmassivenheter2023]. 
These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric predictors will vary by landscape.

@@

The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration [@strackPotentialPeatlandsNatureBased2022].
Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps *before* detailed investigation increases the option space for spatial planning and land management. 
For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work, but by that time, the route of the road is already set. 
With a digital peat depth map, planners could better compare routes and climate mitigation measures would have more leverage.
Therefore, soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations are.

Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at two contrasting sites, with a view towards revising regional and national maps in Norway. 
Specifically, we aim to: (1) quantify the accuracy of predictions from topographic and radiometric variables, and (2) identify key predictive variables.
Reflecting on our results and the need for continued improvement in peat depth maps, we close by discussing implications for digital soil mapping of peat depth.

To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms. 
Where airborne radiometric data have previously been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation [e.g., @keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @siemonAirborneElectromagneticRadiometric2020]. 
Where machine learning algorithms have previously been used on airborne radiometric data, it has been to predict peat extent rather than depth [e.g., @olearyDigitalSoilMapping2022].

@@

At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage and accessibility (Fig. \@ref(fig:sites)b).
The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % monzonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (Geological Survey of Norway, 1:250 000 dataset).
The area within our delineation at Skrimfjella is classified as the landscape type *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
In Norway's AR5 national land cover dataset ["areal resources in scale 1:5000", @ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as 'mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.

@@

We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \@ref(tab:preds)).
Each continuous predictor equates to one variable in the model, while the categorical predictor equates to two variables in the model because its three levels become two indicator variables (one level is used as the reference). 
All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
From the radiometric surveys, we simply used the four variables produced by the Geological Survey of Norway: ground concentration of potassium, thorium, uranium, as well as total count.
From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables [@maxwellLandsurfaceParametersSpatial2022].
The categorical predictor was peat depth class, from a national map dataset.
We chose a spatial resolution of 10 m for our predictors, peat depth sample, and modelling.
We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other hand.
Complete descriptions of predictors follow below.

```{r predictors, echo=FALSE, results='asis'}
predictors <- readr::read_file("tables/predictors.tex")
cat(predictors)
```

@@

The Geological Survey of Norway conducted and processed radiometric surveys over our study areas, as reported in Baranwal et al. [-@baranwalHelicopterborneMagneticElectromagnetic2013] and Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015].
They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the *terra* package (v1.7) in R.

The radiometric surveys were conduced with similar flight parameters (Table \ref{tab:radSurveys}).
Spectrometer count rates were calibrated to known concentrations of potassium, thorium, and uranium in mobile pads.
Data were processed following standard procedures outlined by the International Atomic Energy Association.
Processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
At Ørskogfjellet an additional convolution filter was added to smooth the gridded data.

@@

We used the *terra* package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @dornikOptimalScalingPredictors2022; @newmanAssessingSpatiallyHeterogeneous2023].
We know that peat depth tends to vary at fine scales in Norway, which is why we chose 1 m and 10 m resolutions [@maxwellLandsurfaceParametersSpatial2022].
We also calculated the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
We calculated this index in SAGA GIS [v.9.3.2, Morphometry library, @conradSystemAutomatedGeoscientific2015] with default parameters (initial slope threshold = 16 %, lowness threshold = 0.4, upness threshold = 0.35, slope shape parameter = 4, elevation shape parameter = 3).

@@

The Depth-to-Water index [@murphyMappingWetlandsComparison2007] approximates a location's vertical height above the surface water feature (e.g. stream, lake, or sea) that it is likely to drain towards.
It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature [eq. 5 in @murphyTopographicModellingSoil2009].
We calculated unitless slope from the 1 m DTM using the Whitebox software.
Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthwater2021].
This catchment area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
For each of the seven surface water layers, we derived Depth-to-Water using the *Distance Accumulation* tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.

@@

At Skrimfjella, we used the *eSample* function in the *iSDM* R package (v1.0) to stratify our sample across elevation, slope, and potassium concentration.
This function defines the environmental space as a two-dimensional convex hull around the PCA-ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and *eSample* returned 105 raster cells.

We also measured peatland occurrence (binary variable) in a separate sample at Skrimfjella.
We measured peatland occurrence because the AR5 dataset is known to underestimate peatland extent [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
The 10 m cells were sampled from those outside AR5 mire and with slope < 20°. 
Following the same procedure, *eSample* returned 106 raster cells from this population.

@@

At Ørskogfjellet, we first determined a minimal sample size that would adequately capture the slope and radiometric properties (potassium, thorium, uranium, and total count) of the entire AR5 mire area [@sauretteDivergenceMetricsDetermining2023].
Specifically, we identified an elbow point in a curve of similarity between sample and population [@maloneMethodsImproveUtility2019].
For a sequence of sample sizes (50--500) [ten replicates each, drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback--Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
Then we fitted an asymptotic regression of mean divergence on sample size, and found that the curve reached 95 % of the fitted asymptote at a sample size of 160.

To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the the *fields* package (v14.2).
Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space [@brusSamplingDigitalSoil2019].
This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: slope and four radiometrics.

We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (compared to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.

@@

We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
We dampened the effect of outlying measurements by probing three times at each location [@parryEvaluatingApproachesEstimating2014], at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides. 
We used changes in resistance to indicate the base of the peat column.
Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact, like a buried rock.
Where the peat column was deeper than the extendable probe could be manually inserted and extracted by a pair of operators, we recorded a right-censored result (one at Skrimfjella, five at Ørskogfjellet).
